Phonon polaritonics in two-dimensional materials

### 2019-03-18 paper.tex
just fixing some latex stuff; serious look later in the afternoon

Given these results, we now re-express the conductivity explicitly in terms of the 2D phonon dispersion, and derive a universal form for the conductivity in the local ($q\rightarrow 0$) limit specified in terms of three parameters: the LO phonon frequency at the $\Gamma$ point (i.e., $\omega_{\mathrm{TO}}$), the group velocity of the LO phonon at the $\Gamma$ point, and the damping rate. From Equation~\eqref{eq:zerosofresponse}, we can immediately write the conductivity as
  \begin{equation}
  \sigma(\mathbf{q},\omega) = \frac{-2\iu\epsilon_0\epsilon_{\mathrm{env}}\omega}{q}\frac{\omega^2_{\mathbf{q},\mathrm{LO}}-\omega^2_{\mathrm{TO}}}{\omega^2_{\mathrm{TO}}-\omega^2-\iu\omega\tau^{-1}}.
  \end{equation}
 In this expression, $\omega_\mathrm{L}$, the LO phonon frequency prior to LO-TO splitting, has been re-named as $\omega_{\mathrm{TO}}$, the transverse optical phonon frequency, because in the absence of LO--TO splitting, they are degenerate. The RPA zeros $\omega_{\mathbf{q}}$ have also been renamed as $\omega_{\mathbf{q},\mathrm{LO}}$. This is done in order to make the form of the final results more closely resemble their 3D counterparts, in which the dielectric function is expressed in terms of the TO frequency (see for example Eq. (\ref{eq:3deps})).

@@

The second comment is that much of what has been discussed here has a strong analogy with the theory of optical response in electron gases in 2D, and particularly the relation between plasmons in 2D and 3D. To elaborate on this analogy, we take Equation~\eqref{eq:zerosofresponse} in the case of a two-atom unit cell (such as hBN), and note that the term in the sum over Born charges can be written as $\Big|\hat{\mathbf{q}}\cdot \sum_\kappa \mathbf{Z}_\kappa\boldsymbol{\eta}_\kappa\Big|^2 \equiv \frac{Q_*^2}{M_*}$, with $Q_*$ being an effective charge and $M_*$ being an effective mass. Then, the LO--TO splitting can be written as $\omega^2_{\mathbf{q}} - \omega^2_{\mathbf{q},\mathrm{TO}} = \frac{Q_*^2}{2\epsilon_0\epsilon_{\mathrm{env}} M_*}q$. Now we note that the RHS is exactly the squared-frequency $\omega^2_{\mathbf{q}\mathrm{p}}$ for a plasma oscillation in a 2D gas of charged particles with charge $Q_*$ and mass $M_*$. To connect to LO-TO splitting in phonons, this squared frequency $\omega^2_{\mathbf{q}\mathrm{p}}$ can be thought of as the ``LP--TP'' splitting between longitudinal and transverse plasma oscillations. Of course, there are no transverse plasma oscillations due to the structure-less nature of the electron, and so ``$\omega_{\mathrm{TP}}$''  should be considered equal to zero. In the three-dimensional plasmon case, ``$\omega_{\mathrm{TP}} = 0$'', but the difference between the squared longitudinal and transverse plasma oscillation frequencies at zero-wavevector is non-zero and given by $\omega_{\mathrm{p}}^2$. In other words, the plasma frequency in electron gases is analogous to the LO--TO splitting in polar materials. The change in the dispersion of 3D versus 2D plasmons, like 3D versus 2D LO phonons, arises from the change in dimensionality of the Coulomb interaction from 3D to 2D.  This analogy between phononic and plasmonic behavior as a function of dimension is illustrated schematically in Figures~\ref{fig:1}(b,c) in order to help unify the understanding of 2D plasmonics and phononics. These considerations should also extend to the one-dimensional case.

The third comment is that the considerations of this section can be extended to few-layer systems, such as hBN bilayers, which may still be considered 2D for polariton wavelengths large compared to the bilayer thickness. The considerations may also be extended to 1D polar materials, such as BN nanotubes. For bilayers, as was discussed in Ref.~ \cite{sohier2017breakdown}, one may describe the LO-TO splitting in terms of Eq. ~\ref{eq:zerosofresponse}, with the 2D Coulomb interaction, but with eigendisplacements appropriate to the few layer system. For a small number of layers $N$, this should enhance the LO-TO splitting by a factor of $N$. For 1D systems, beyond using the eigendisplacements appropriate to the 1D system, one must also use the 1D Coulomb interaction, cutoff by a length $R$ which for a nanotube, would be its radius. In that case, the 1D Coulomb interaction is $\frac{e^2K_0(qr)}{2\pi\epsilon_0}$, where $K_0$ is a modified Bessel function of the second kind and of order zero. Thus, the $q$-dependence of the LO-TO splitting is $q^2K_0(qR)$. This is in close analogy with the relation for the plasma frequency of 1D plasmons, as expected from the considerations of the previous paragraph. As regards the optical properties, the optical conductivity is still governed by Eq.~\ref{eq:2dsusceptibility}.

%\textcolor{blue}{The third comment is that the considerations of this section can be extended to few-layer systems, such as hBN bilayers, which may still be considered 2D for polariton wavelengths large compared to the bilayer thickness. The considerations may also be extended to 1D polar materials, such as BN nanotubes. For bilayers, as was discussed in Ref.~ \cite{sohier2017breakdown}, one may describe the LO-TO splitting in terms of Eq. ~\ref{eq:zerosofresponse}, with the 2D Coulomb interaction, but with eigendisplacements appropriate to the few layer system. For a small number of layers $N$, this should enhance the LO-TO splitting by a factor of $N$. For 1D systems, beyond using the eigendisplacements appropriate to the 1D system, one must also use the 1D Coulomb interaction, cutoff by a length $R$ which for a nanotube, would be its radius. In that case, the 1D Coulomb interaction is $\frac{e^2K_0(qr)}{2\pi\epsilon_0}$, where $K_0$ is a modified Bessel function of the second kind and of order zero. Thus, the $q$-dependence of the LO-TO splitting is $q^2K_0(qR)$. This is in close analogy with the relation for the plasma frequency of 1D plasmons, as expected from the considerations of the previous paragraph. As regards the optical properties, the optical conductivity is still governed by Eq.~\ref{eq:2dsusceptibility}.}

@@

In the rest of this section, we discuss the relation between the dispersion of an hBN monolayer versus the atomically-thick limit of a thin film of a material with hBN's bulk dielectric function. To aid this discussion, in Figure~\ref{fig:2}(c), we show the dispersion relation of thin films of bulk hBN with film thicknesses of 1, 2, 4, and 8 times the interlayer spacing of bulk hBN, which is roughly \SI{0.33}{\nm}. For these plots, we take hBN to be cleaved such that the optical axis is perpendicular to the plane of the film. The components of the bulk permittivity perpendicular and parallel to the c-axis ($\epsilon_{\perp}$ and $\epsilon_{\parallel}$, respectively; indexed by $\alpha\in\{\perp,\parallel\}$ below) are then given by
\begin{equation}
	\epsilon_{\alpha}(\omega) = \epsilon_{\infty,\alpha}\Bigg(1 + \frac{\omega^2_{\mathrm{LO},\alpha}-\omega^2_{\mathrm{TO},\alpha}}{\omega^2_{\mathrm{TO},\alpha}-\omega^2} \Bigg),
	\label{eq:3deps}
\end{equation}
with $\epsilon_{\infty,\parallel} = \num{2.95}$, $\omega_{\mathrm{TO},\parallel} = \SI{760}{\per\cm}$, and $\omega_{\mathrm{LO},\parallel} = \SI{830}{\per\cm}$; and $\epsilon_{\infty,\perp} = \num{4.87}$, $\omega_{\mathrm{TO},\perp} = \SI{1360}{\per\cm}$, and $\omega_{\mathrm{LO},\perp} = \SI{1614}{\per\cm}$\cite{caldwell2014sub,dai2014tunable}.
Losses are ignored in this discussion altogether. 
In the range between $\omega_{\mathrm{TO}, \alpha}$ and $\omega_{\mathrm{LO}, \alpha}$, the corresponding component of the permittivity is negative, while the other component is positive. This hyperbolicity leads to a dispersion for hBN thin films that have multiple branches at a given frequency, as can be seen in Figure~\ref{fig:2}(c).  This trend persists even when the thickness of the bulk is taken down to a single layer, albeit being pushed to high wavevectors. This is in contrast with the true monolayer, in which there is only one LO phonon mode. Given that the phonon polariton of 2D is the LO phonon, there can be only one branch of the dispersion. The bulk dielectric function of Eq.~\ref{eq:3deps} fails to capture this lack of higher-order modes, and thus hyperbolicity. For a bilayer however, there is a higher-order LO mode, associated with out-of-phase oscillation between the two layers, and thus one expects to see a higher-order mode. However, it is expected challenging to observe such a mode, as can be anticipated from the blue higher-order mode in Fig.~\ref{fig:2}(c), because even for a polariton wavelength of 12 nm, which is very short, the polariton frequency is only a few cm$^{-1}$ above the TO frequency, within the damping linewidth of the TO frequency. %\textcolor{blue}{The bulk dielectric function of Eq.~\ref{eq:3deps} fails to capture this lack of higher-order modes, and thus hyperbolicity. For a bilayer however, there is a higher-order LO mode, associated with out-of-phase oscillation between the two layers, and thus one expects to see a higher-order mode. However, it is expected challenging to observe such a mode, as can be anticipated from the blue higher-order mode in Fig.~\ref{fig:2}(c), because even for a polariton wavelength of 12 nm, which is very short, the polariton frequency is only a few cm$^{-1}$ above the TO frequency, within the damping linewidth of the TO frequency.}

Another anomaly between 2D and 3D is that the dispersion relation of the 2D LO phonon is non-monotonic, and the frequency starts to decrease for sufficiently large wavevector. This can never happen by extrapolating to a monolayer a dielectric function which has finite LO--TO splitting. This turn-back point corresponds to a frequency of around \SI{1500}{\per\cm} and a phonon wavelength of less than \SI{1}{\nm}. Thus, this deviation is challenging to probe in typical optical experiments. However it means that for the monolayer, the range of accessible phonon polaritonic frequencies for optical applications is reduced compared to thick films. Finally, while we have focused on anomalies between the atomically-thin limit of bulk and a true monolayer, a comparison of the monolayer with the fundamental phonon polariton mode of the one-atom-thick thin film suggests that we have a reasonable qualitative understanding of the monolayer dispersion from this limit. This qualitative similarity is to be somewhat expected, as in many cases, a 2D layer can be considered as an very thin film of bulk, provided that the microscopic properties of the bulk and monolayer are the same. In fact, if we take an thin film of thickness $t$ of hBN of dielectric function given by Eq.~\ref{eq:3deps}, and focus only on the in-plane response, then there is a mathematical condition for the thin film to exactly mirror the bulk. It is: 
\begin{equation}
(\omega^{2 (\mathrm{bulk})}_{\mathrm{LO}} - \omega^{2 (\mathrm{bulk})}_{\mathrm{TO}})t = 4\omega^{(\mathrm{mono})}_{\mathrm{TO}}v^{(\mathrm{mono})}_g,
\end{equation}
where "bulk" denotes a property of the bulk dielectric function and "mono" denotes that of the monolayer conductivity.

>>>>

Given these results, we now re-express the conductivity explicitly in terms of the 2D phonon dispersion, and derive a universal form for the conductivity in the local ($q\rightarrow 0$) limit specified in terms of three parameters: the LO phonon frequency at the $\Gamma$ point (i.e., $\omega_{\mathrm{TO}}$), the group velocity of the LO phonon at the $\Gamma$ point, and the damping rate. From Equation~\eqref{eq:zerosofresponse}, we can immediately write the conductivity as
  \begin{equation}
  \sigma(\mathbf{q},\omega) = \frac{-2\iu\epsilon_0\epsilon_{\mathrm{env}}\omega}{q}\frac{\omega^2_{\mathbf{q},\mathrm{LO}}-\omega^2_{\mathrm{TO}}}{\omega^2_{\mathrm{TO}}-\omega^2-\iu\omega\tau^{-1}}.
  \end{equation}
 In this expression, $\omega_\mathrm{L}$, the LO phonon frequency prior to LO--TO splitting, has been re-named as $\omega_{\mathrm{TO}}$, the transverse optical phonon frequency, because in the absence of LO--TO splitting, they are degenerate. The RPA zeros $\omega_{\mathbf{q}}$ have also been renamed as $\omega_{\mathbf{q},\mathrm{LO}}$. This is done in order to make the form of the final results more closely resemble their 3D counterparts, in which the dielectric function is expressed in terms of the TO frequency (see for example Eq. (\ref{eq:3deps})).

@@

The second comment is that much of what has been discussed here has a strong analogy with the theory of optical response in electron gases in 2D, and particularly the relation between plasmons in 2D and 3D. To elaborate on this analogy, we take Equation~\eqref{eq:zerosofresponse} in the case of a two-atom unit cell (such as hBN), and note that the term in the sum over Born charges can be written as $\Big|\hat{\mathbf{q}}\cdot \sum_\kappa \mathbf{Z}_\kappa\boldsymbol{\eta}_\kappa\Big|^2 \equiv \frac{Q_*^2}{M_*}$, with $Q_*$ being an effective charge and $M_*$ being an effective mass. Then, the LO--TO splitting can be written as $\omega^2_{\mathbf{q}} - \omega^2_{\mathbf{q},\mathrm{TO}} = \frac{Q_*^2}{2\epsilon_0\epsilon_{\mathrm{env}} M_*}q$. Now we note that the RHS is exactly the squared-frequency $\omega^2_{\mathbf{q}\mathrm{p}}$ for a plasma oscillation in a 2D gas of charged particles with charge $Q_*$ and mass $M_*$. To connect to LO--TO splitting in phonons, this squared frequency $\omega^2_{\mathbf{q}\mathrm{p}}$ can be thought of as the ``LP--TP'' splitting between longitudinal and transverse plasma oscillations. Of course, there are no transverse plasma oscillations due to the structure-less nature of the electron, and so ``$\omega_{\mathrm{TP}}$''  should be considered equal to zero. In the three-dimensional plasmon case, ``$\omega_{\mathrm{TP}} = 0$'', but the difference between the squared longitudinal and transverse plasma oscillation frequencies at zero-wavevector is non-zero and given by $\omega_{\mathrm{p}}^2$. In other words, the plasma frequency in electron gases is analogous to the LO--TO splitting in polar materials. The change in the dispersion of 3D versus 2D plasmons, like 3D versus 2D LO phonons, arises from the change in dimensionality of the Coulomb interaction from 3D to 2D.  This analogy between phononic and plasmonic behavior as a function of dimension is illustrated schematically in Figures~\ref{fig:1}(b,c) in order to help unify the understanding of 2D plasmonics and phononics. These considerations should also extend to the one-dimensional case.

The third comment is that the considerations of this section can be extended to few-layer systems, such as hBN bilayers, which may still be considered 2D for polariton wavelengths large compared to the bilayer thickness. The considerations may also be extended to 1D polar materials, such as BN nanotubes. For bilayers, as was discussed in Ref.~\citenum{sohier2017breakdown}, one may describe the LO--TO splitting in terms of Eq.~\eqref{eq:zerosofresponse}, with the 2D Coulomb interaction, but with eigendisplacements appropriate to the few layer system. For a small number of layers $N$, this should enhance the LO--TO splitting by a factor of $N$. For 1D systems, beyond using the eigendisplacements appropriate to the 1D system, one must also use the 1D Coulomb interaction, cutoff by a length $R$ which for a nanotube, would be its radius. In that case, the 1D Coulomb interaction is $e^2K_0(qr)/2\pi\epsilon_0$, where $K_0$ is the (zeroth order) modified Bessel function of the second kind. Thus, the $q$-dependence of the LO--TO splitting is $q^2K_0(qR)$. This is in close analogy with the relation for the plasma frequency of 1D plasmons, as expected from the considerations of the previous paragraph. As regards the optical properties, the optical conductivity is still governed by Eq.~\eqref{eq:2dsusceptibility}.

%\textcolor{blue}{The third comment is that the considerations of this section can be extended to few-layer systems, such as hBN bilayers, which may still be considered 2D for polariton wavelengths large compared to the bilayer thickness. The considerations may also be extended to 1D polar materials, such as BN nanotubes. For bilayers, as was discussed in Ref.~ \cite{sohier2017breakdown}, one may describe the LO--TO splitting in terms of Eq. ~\ref{eq:zerosofresponse}, with the 2D Coulomb interaction, but with eigendisplacements appropriate to the few layer system. For a small number of layers $N$, this should enhance the LO--TO splitting by a factor of $N$. For 1D systems, beyond using the eigendisplacements appropriate to the 1D system, one must also use the 1D Coulomb interaction, cutoff by a length $R$ which for a nanotube, would be its radius. In that case, the 1D Coulomb interaction is $\frac{e^2K_0(qr)}{2\pi\epsilon_0}$, where $K_0$ is a modified Bessel function of the second kind and of order zero. Thus, the $q$-dependence of the LO--TO splitting is $q^2K_0(qR)$. This is in close analogy with the relation for the plasma frequency of 1D plasmons, as expected from the considerations of the previous paragraph. As regards the optical properties, the optical conductivity is still governed by Eq.~\ref{eq:2dsusceptibility}.}

@@

In the rest of this section, we discuss the relation between the dispersion of an hBN monolayer versus the atomically-thick limit of a thin film of a material with hBN's bulk dielectric function. To aid this discussion, in Figure~\ref{fig:2}(c), we show the dispersion relation of thin films of bulk hBN with film thicknesses of 1, 2, 4, and 8 times the interlayer spacing of bulk hBN, which is roughly \SI{0.33}{\nm}. For these plots, we take hBN to be cleaved such that the optical axis is perpendicular to the plane of the film. The components of the bulk permittivity perpendicular and parallel to the c-axis ($\epsilon_{\perp}$ and $\epsilon_{\parallel}$, respectively; indexed by $\alpha\in\{\perp,\parallel\}$ below) are then given by
\begin{equation}
	\epsilon_{\alpha}(\omega) = \epsilon_{\infty,\alpha}\Bigg(1 + \frac{\omega^2_{\mathrm{LO},\alpha}-\omega^2_{\mathrm{TO},\alpha}}{\omega^2_{\mathrm{TO},\alpha}-\omega^2} \Bigg),
	\label{eq:3deps}
\end{equation}
with $\epsilon_{\infty,\parallel} = \num{2.95}$, $\omega_{\mathrm{TO},\parallel} = \SI{760}{\per\cm}$, and $\omega_{\mathrm{LO},\parallel} = \SI{830}{\per\cm}$; and $\epsilon_{\infty,\perp} = \num{4.87}$, $\omega_{\mathrm{TO},\perp} = \SI{1360}{\per\cm}$, and $\omega_{\mathrm{LO},\perp} = \SI{1614}{\per\cm}$\cite{caldwell2014sub,dai2014tunable}.
Losses are ignored in this discussion altogether. 
In the range between $\omega_{\mathrm{TO}, \alpha}$ and $\omega_{\mathrm{LO}, \alpha}$, the corresponding component of the permittivity is negative, while the other component is positive. This hyperbolicity leads to a dispersion for hBN thin films that have multiple branches at a given frequency, as can be seen in Figure~\ref{fig:2}(c).  This trend persists even when the thickness of the bulk is taken down to a single layer, albeit being pushed to high wavevectors. This is in contrast with the true monolayer, in which there is only one LO phonon mode. Given that the phonon polariton of 2D is the LO phonon, there can be only one branch of the dispersion. The bulk dielectric function of Eq.~\eqref{eq:3deps} fails to capture this lack of higher-order modes, and thus hyperbolicity. For a bilayer however, there is a higher-order LO mode, associated with out-of-phase oscillation between the two layers, and thus one expects to see a higher-order mode. However, it is expected challenging to observe such a mode, as can be anticipated from the blue higher-order mode in Fig.~\ref{fig:2}(c), because even for a polariton wavelength of \SI{12}{\nm}, which is very short, the polariton frequency is only a few \si{\per\cm} above the TO frequency, within the damping linewidth of the TO frequency. %\textcolor{blue}{The bulk dielectric function of Eq.~\ref{eq:3deps} fails to capture this lack of higher-order modes, and thus hyperbolicity. For a bilayer however, there is a higher-order LO mode, associated with out-of-phase oscillation between the two layers, and thus one expects to see a higher-order mode. However, it is expected challenging to observe such a mode, as can be anticipated from the blue higher-order mode in Fig.~\ref{fig:2}(c), because even for a polariton wavelength of 12 nm, which is very short, the polariton frequency is only a few cm$^{-1}$ above the TO frequency, within the damping linewidth of the TO frequency.}

Another anomaly between 2D and 3D is that the dispersion relation of the 2D LO phonon is non-monotonic, and the frequency starts to decrease for sufficiently large wavevector. This can never happen by extrapolating to a monolayer a dielectric function which has finite LO--TO splitting. This turn-back point corresponds to a frequency of around \SI{1500}{\per\cm} and a phonon wavelength of less than \SI{1}{\nm}. Thus, this deviation is challenging to probe in typical optical experiments. However it means that for the monolayer, the range of accessible phonon polaritonic frequencies for optical applications is reduced compared to thick films. Finally, while we have focused on anomalies between the atomically-thin limit of bulk and a true monolayer, a comparison of the monolayer with the fundamental phonon polariton mode of the one-atom-thick thin film suggests that we have a reasonable qualitative understanding of the monolayer dispersion from this limit. This qualitative similarity is to be somewhat expected, as in many cases, a 2D layer can be considered as an very thin film of bulk, provided that the microscopic properties of the bulk and monolayer are the same. In fact, if we take an thin film of thickness $t$ of hBN of dielectric function given by Eq.~\eqref{eq:3deps}, and focus only on the in-plane response, then there is a mathematical condition for the thin film to exactly mirror the bulk. It is: 
\begin{equation}
(\omega^{2 (\mathrm{bulk})}_{\mathrm{LO}} - \omega^{2 (\mathrm{bulk})}_{\mathrm{TO}})t = 4\omega^{(\mathrm{mono})}_{\mathrm{TO}}v^{(\mathrm{mono})}_g,
\end{equation}
where `bulk' denotes a property of the bulk dielectric function and `mono' denotes that of the monolayer conductivity.
